Gap formation and stability in non-isothermal protoplanetary discs

### 2015-02-15 paper/response.txt

- §§4 and 5: would the results of a simulation with a very short
  cooling time (beta << 1) be the same as those of a simulation with a
  locally isothermal equation of state and same unperturbed sound
  speed? 
  ********************************************************************
  We have performed addtional (low resolution) simulations with
  beta=0.01 to compare with a locally isothermal disks. We find a
  smaller linear growth rate in the locally isothermal case.

  ********************************************************************

- §4.4: why does the vortex’s perturbed density decrease on the long
  term? I suppose a related question is: what controls the long-term
  evolution of the vortex’s potential vorticity or entropy-modified
  potential vorticity? (whichever quantity is relevant). 
  ********************************************************************
  In theory a vortex should live forever in 2D inviscid flow. In
  practice, however, there is numerical viscosity which we suspect is
  responsible for the long term decay. As the vortex weakens
  (elongates) its area increases and its surface density drops.

  viscosity and elongation of vortex

  ask robert: does gap shape change significantly?

  ********************************************************************

- §5.2-5.3: here again I’d like to have more details on the vortex
        evolution and decay: 	
        . Do the authors see the vortex migrate? I did a simulation
        with the authors’ setup and a rather long cooling time
        (beta_tilde ~ 8) where I see the vortex moving from ~6.5 rh to
        ~5.3 rh relative to the planet’s orbital radius before the
        vortex starts decaying (rh = Hill radius). The time when the
        vortex starts decaying actually corresponds to a rather sudden
        increase in the planet-vortex distance, which goes up to ~6.8
        rh until the vortex dies off completely. If the authors get a
        similar migration behaviour in their simulations, maybe this
        could give further insight into the vortex evolution (vortex
        getting stronger during the quasi steady state before fading
        away). The connection between migration and the appearance of
        spiral waves launched by the vortex could also be made. 
	********************************************************************
        We thank the reviewer for raising the issue of vortex
        migration. We have re-examined our simulation results. We
        indeed observe vortex migration but our result is somewhat
        different to that of the reviewer.

 	We do not see significant vortex migration during quasi-steady
 	state. This may be due to the fact that the
 	vortex forms at a surface density maximum, which is able to
 	trap vortices (Paardekooper et al 2010).

	Instead, we observe the vortex begins to migrate inwards
	with the appearence of the spiral wakes (cf. planetary
	migration). However, we find the vortex eventually migrates
	outward to about 6.5rh before complete decay.

	We remark that for a vortex formed at a surface density
 	maximum (e.g. from the RWI), that subsequent perturbations may
 	lead to inwards or outwards vortex migration, depending on
 	what side of the maximum the vortex is displaced to
 	(Paardekooper et al 2010).

	Although we do not understand the discrepency between our
	results and the reviewer's, we do agree with the reviewer
	that vortex migration probably plays a role in determining
	the overall vortex lifetime (it may affect the duration of the
	quasi-steady state and/or the decay timescale). We now mention
	this in the discussions and caveats.

        ********************************************************************

        . During the quasi steady state the vortex grows as a result
        of continuous vortensity generation upon crossing the planet’s
        wake. What about possible turbulent diffusion of potential
        vorticity if the planet forms in a region of the disc with
        non-negligible ‘turbulent viscosity’? Have the authors done
        simulations with a non vanishingly small viscosity? 
        ********************************************************************
	Since a low-viscosity disc is needed for the RWI to operate,
        we have chosen to focus on invisicd simulations. This also
        helps to minimize vortex decay directly due to a physical
        viscosity, so that we can examine the influence of
        thermodynamics.

	Thus, we expect that vortex lifetimes in realistic viscous discs
	to be shorter than that seen in our 
	simulations. We mention this in the discussions.

        ********************************************************************
        . I was a bit surprised that the RWI couldn’t be launched
        again after the vortex had decayed, and I’d like the authors
        to be more specific about why it is so. About the vortex decay
        providing an effective alpha viscosity~ 1d-2: isn’t this a
        time-varying quantity, decreasing as the vortex is dying? If
        so is it really the reason why the RWI isn’t launched again?
        What about the disc’s aspect ratio at the outer gap edge after
        the vortex’s full decay?     
        ********************************************************************
	The quoted value of 0.01 is a typical viscosity associated
        with the vortex, averaged over radius and time.

	By the time the vortex dies out that there is
	no longer a `vortex viscosity', but the outer gap edge
	(surface density bump) has already been smoothed out, and the
	site is no longer suitable for the RWI. Note that from
	Fig. 11, smoothing out the surface density bump puts material
	into the outer disc.

	In order to re-launch the RWI, the surface density bump needs
	to reform. However, this seems difficult as the planet already
	resides in a deep gap with little material to clear out
	(c.f. the original surface density bump resulted from
	gap-opening: the planet pushes material to either side of its
	orbit).

        ********************************************************************

- §6.1: two caveats / outlooks could be added: (i) having a more
  realistic cooling model (same caveat as for numerical simulations of
  disc fragmentation by the gravitational instability…) and (ii)
  considering planet migration. 
  ********************************************************************
  These are now added. 
  ********************************************************************
- §2: given the very low shear viscosity used in this work, I would
  emphasize straight after eqs. (2) and (3) that f_v and H include the
  contribution of a (Von Neumann-Ritchmyer) artificial bulk viscosity
  to handle shocks. 
  ********************************************************************
  These comments are now added. 
  ********************************************************************
- §4.2: the minimum value of 2d-4 for kappa^2 + N^2 should rather be
  expressed in units of Omega^2 at the planet’s orbital radius. Same
  advice for N, for which the value of 1d-5 didn’t really speak to me
  until I realised that the planet’s orbital radius wasn’t the code’s
  unit of length. By the way, where is N evaluated? 
  ********************************************************************
  The frequencies are now expressed in units of the orbital frequency
  at the planet's orbital radius. 
  ********************************************************************
- Fig. 6: the colour scale suggests that there is a discontinuous jump
  in perturbed density at orbital radii R > 40 r_in, where the density
  seems to take very small values… why is that? 
  ********************************************************************
  This is due to the open boundary conditions adopted at the outer
  boundary, which leads to mass loss. We now mention this in the
  caption of Fig. 6. 
  ********************************************************************

### 2015-02-15 paper/model.tex
revision in the works

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum 
equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$ .

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disk.

>>>>

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
which includes artificial viscosity to handle shocks, and a
Navier-Stokes viscosity } whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating {\bf(from both physical and
  artificial viscosity)} and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disk.

### 2015-02-15 paper/results.tex
revision in the works

\begin{table}
  \centering
  \caption{Dominant mode and growth rates for
    $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
    values during `planet-off' simulations \label{modetable}} 
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      6 & 7.3 \\
      7 & 7.8 \\
      8 & 7.9 \\
      9 & 7.9 \\
      10 & 6.8 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      3 & 2.0 \\
      4 & 2.2 \\
      5 & 2.3 \\
      6 & 1.6 \\
      7 & 1.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      1 & 1.1 \\
      2 & 1.6 \\
      3 & 1.7 \\
      4 & 1.2 \\
      5 & 0.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
\end{table}

@@

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterise these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

@@

Table \ref{modetable} show that as
$\tilde{\beta}$ is increased from $ 0.1\rightarrow10$ the dominant
azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
respective growth rate decreases from $ \gamma/\Omega(r_p)=0.079
\rightarrow 0.017$. However, despite two orders of magnitude increase in the
cooling time, the instability remains dynamical with characteristic  growth time
$\lesssim 10P_0$. Snapshots of the instability in for  
the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}. {\bf Isothermal simulations with set sound speed where also analyzed showing growth rates $\sim 5 \times 10^2 / \Omega(r_p)$. In comparison nearly isothermal simulation with $\tilde\beta=0.01$ had growth rates $\sim 8\times 10^2 / \Omega(r_p)$}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the initially formed vortex was 
found to decrease with increased cooling rate. The vortices simply decay on a
timescale of $O(10^3)$ orbits with faster decay for stronger vortices
(which are obtained with faster cooling rates). We will see in the
next section that this behaviour is very different to when the planet
potential is kept on.

@@

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex at the outer gap edge. 
    The plots for other values of the cooling time
    $\tilde{\beta}$ are similar.
    {\bf The decrease in density pertibation at $r>40r_{in}$ arises from the open
    boundry condition.}
    \label{Vortex2D}} 
\end{figure}

@@

@@

{\bf Radial locations of the vortices were found to migrate over their lifetime. During the quasi-steady state the vortices were found to orbit $\sim 6.5r_h$ from the planet radius. Simultaneously with the appearence of wakes, the vortices were found to move towards $\sim 5.3r_h$. This is a result of the the spiral waves transporting angular momentum to the disk (Is this true? Shocks on the inner disk seem stronger so I would think it would migrate outward). After the vortex reaches marginal amplitudes $<1 \Delta\Sigma/\Sigma$ the vortices were found to migrate back out towards $\sim 6.5 r_h$ from the planet location }

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
$w_s\simeq0.1$ (code units) is a typical length scale of 
the surface density variation across the wake.   
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 0.6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
begins decreasing to zero
%{\bf :check}.

@@

A plot of the gradient parameter over time
for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
During vortex decay, the outer gap edge is 
drastically smoothed out, changing from a value of $\delta\Sigma=1.2$ during
quasi-steady state to $0.4$ after dissipation. 
%and we find that the gap edge is smoothed during this
%process. 
This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, so the condition for the RWI
becomes less favourable, which may explain why vortices do not
reform again (at least within the simulation timescale). 
%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

>>>>

\begin{table}
  \centering
  \caption{Dominant mode and growth rates for
    $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
    values during `planet-off' simulations \label{modetable}} 
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
      \toprule
      $m$ & $10^2\sigma/\Omega(r_p)$ \\
      \midrule
      6 & 7.3 \\
      7 & 7.8 \\
      8 & 7.9 \\
      9 & 7.9 \\
      10 & 6.8 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
      \toprule
      $m$ & $10^2\sigma/\Omega(r_p)$ \\
      \midrule
      3 & 2.0 \\
      4 & 2.2 \\
      5 & 2.3 \\
      6 & 1.6 \\
      7 & 1.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
      \toprule
      $m$ & $10^2\sigma/\Omega(r_p)$ \\
      \midrule
      1 & 1.1 \\
      2 & 1.6 \\
      3 & 1.7 \\
      4 & 1.2 \\
      5 & 0.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
\end{table}

@@

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterise these
modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

@@

Table \ref{modetable} show that as
$\tilde{\beta}$ is increased from $ 0.1\rightarrow10$ the dominant
azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
respective growth rate decreases from $ \gamma/\Omega(r_p)=0.079
\rightarrow 0.017$. However, despite two orders of magnitude increase in the
cooling time, the instability remains dynamical with characteristic  growth time
$\lesssim 10P_0$. Snapshots of the instability in for  
the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.

{\bf We also simulated a locally isothermal disk where the sound-speed is
  held fixed, which yield a most unstable growth rate
  $\sigma/Omega(r_p)\simeq 0.05$, compared with a value of $0.08$ for
  a case with $\tilde{\beta}=0.01$.}   
%Isothermal simulations with set sound speed where also analyzed
%  showing growth rates $\sim 5 \times 10^{-2} / \Omega(r_p)$. }
%In comparison nearly isothermal simulation with $\tilde\beta=0.01$ had growth rates $\sim 8\times 10^52 / \Omega(r_p)$}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the initially formed vortex was 
found to decrease with increased cooling rate. The vortices simply decay on a
timescale of $O(10^3)$ orbits with faster decay for stronger vortices
(which are obtained with faster cooling rates). {\bf This decay is 
  probably due to numerical viscosity. During the slow decay the
  vortex elongates (weakens) while its radial width remains $O(H)$, so
  its surface density decreases. (true?)} 
We will see in the
next section that this behaviour {\bf after linear growth} is very
different to when the planet potential is kept on.

@@

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex at the outer gap edge. The plot for other values of the cooling time
    $\tilde{\beta}$ are similar. {\bf The decrease in the surface
      density near} $r\sim 40 r_\mathrm{in}$ {\bf arises from mass loss due to the open
      boundary condition imposed at} $r_\mathrm{out}$.
    \label{Vortex2D} }
\end{figure}

@@

{\bf For comparison we also ran a locally isothermal simulation to
  compare with the case with rapid cooling ($\tilde{\beta}=0.01$), but
  did not observe significant vortex decay within our simulation
  timescale in the former case. This suggests there could be
  fundamental differences between the two cases.}

@@

{\bf During the quasi-steady state the vortex 
  orbits at $r\sim r_p+6r_h$. We do not see significant
  vortex migration at this stage (true?), since the vortex is located at a
  surface density maximum, which can trap vortices
  \citep{paardekooper10}. However, simultaneous with the
  appearence of the wakes, we observe the vortex begins to migrate 
  inwards to $r\sim r_p+5r_h$. 
  % This is a result of the the spiral waves transporting
  % angular momentum to the disk (Is this true? Shocks on the inner disk
  % seem stronger so I would think it would migrate outward). 
  }

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
$w_s\simeq0.1$ (code units) is a typical length scale of 
the surface density variation across the wake.   
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 0.6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
begins decreasing to zero. {\bf After the 
  vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
  migrates out to $r\sim r_p+ 6.5 r_h$. 
}
%{\bf :check}.

@@

A plot of the gradient parameter over time
for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
During vortex decay, the outer gap edge is 
drastically smoothed out, changing from a value of $\delta\Sigma=1.2$ during
quasi-steady state to $0.4$ after dissipation. 
%and we find that the gap edge is smoothed during this
%process.

This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet {\bf, and smooths out the outer surface density bump,    
} so the condition for the RWI 
becomes less favourable. {\bf In
  order to re-launch the RWI, the surface density bump should
  reform. However, this is difficult as there is no more material
  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}).} {\bf This may} may explain
why vortices do  not reform again (at least within the simulation timescale). 
{\bf what is aspect-ratio at outer gap edge after full decay?}

%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

### 2015-02-15 paper/summary.tex
revision in the works

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }
\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of
non-isothermal disc-planet interaction.  
Our simulations were customised to examine the effect of a finite 
cooling time on the stability of gaps  
opened by giant planets to the so-called vortex or Rossby wave
instability. To do so, we  
included an energy equation with a cooling term that restores the 
disc temperature to its initial profile on a characteristic timescale
$t_c$. We studied the evolution of the gap stability as a function of 
$t_c$. This is a natural extension to previous studies of on gap
stability, which employ locally or strictly isothermal equations of
state.

@@

The duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. The timescale for the vortex to
decay, after reaching maximum amplitude, decreases monotonically with
increasing cooling time. The sum of these two timescales results in a
double-peaked vortex lifetime as a function of cooling
timescale. 
%technically we should probably say there is an optimum cooling time
%to maximise the quasi-steady state

@@

>>>>

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }
\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of
non-isothermal disc-planet interaction.  
Our simulations were customised to examine the effect of a finite 
cooling time on the stability of gaps  
opened by giant planets to the so-called vortex or Rossby wave
instability. To do so, we  
included an energy equation with a cooling term that restores the 
disc temperature to its initial profile on a characteristic timescale
$t_c$. We studied the evolution of the gap stability as a function of 
$t_c$. This is a natural extension to previous studies of on gap
stability, which employ locally or strictly isothermal equations of
state. {\bf We considered the inviscid limit which favors the RWI
  \citep{li09,fu14}, so that the vortex lifetimes observed in our
  simulations are likely longer than in realistic disks with non-zero
  viscosity. }

@@

The duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. The timescale for the vortex to
decay, after reaching maximum amplitude, decreases monotonically with
increasing cooling time. {\bf We observe vortex migration during this
  decay, which may contribute to this decay timescale.} The sum of
these two timescales results in a double-peaked vortex lifetime as a
function of cooling timescale. 
%technically we should probably say there is an optimum cooling time
%to maximise the quasi-steady state

@@

{\bf
  \emph{Orbital migration.} We have held the planet fixed on a circular
  orbit. However, gap-edge vortices are known to exert
  significant, oscillatory torques on the planet \citep{li09}
  which can lead to complex orbital migration. This will likely affect
  vortex lifetimes as it may alter the planet-vortex separation, as
  well as leading to direct vortex-planet interactions
  \citep{lin10,ataiee14}. Thus, 
  future simulations should allow the planet to freely
  migrate. Similarly, the role of vortex migration on its lifetime
  should be clarified. 
}

{\bf 
  \emph{Cooling model.} Our prescription for the disk 
  heating/cooling is convenient to probe the full range of
  thermodynamic response of the disk. However, in order to calculate 
  vortex lifetimes in actual 
  protoplanetary disks, a more realistic thermodynamics treatment ,
  e.g. based on radiative cooling, should be used in future work.   
}
